Biogeography and Environmental Conditions Shape Phage and Bacteria Interaction Networks Across the Healthy Human Microbiome

### 2017-03-30 doc/manuscript.Rmd
More updates.

# Abstract
Viruses and bacteria are critical components of the human microbiome and play important roles in health and disease. Previous work has relied on studying bacteria and phages in isolation, reducing them to two separate communities. These approaches have not captured how these communities interact to share information (e.g. horizontal gene transfer) and maintain stability. We developed and implemented a network-based analytical approach to provide an initial understanding of phage-bacteria network diversity throughout the human body. We accomplished this by building a machine learning algorithm to predict which phages infected which bacteria in a given microbiome. This algorithm was applied to paired viral and bacterial metagenomic sequence sets from three previously published human cohorts. We organized the predicted interactions into networks that allowed us to evaluate the diversity of phage-bacteria connectedness across the human body. We focused on establishing a foundational understanding of microbiome network structures across the human body. We found that gut and skin network structure was person-specific, and was not conserved among cohabitating family members. High fat diets and obesity were associated with less connected networks. There were significant differences in network structure between skin sites, with those exposed to the external environment being less connected and more prone to instability. Altogether this study characterizes the diversity of microbiome networks across the human body, and provides a baseline for future studies to investigate the role of ecological networks in disease states.

@@

Previous work has relied on studying bacterial and phage communities in isolation, reducing them to two separate communities. In reality, bacteria and phage communities are dynamic and complex, acting in concert to share genetic information and to maintain stable ecosystems, and removal of members can disrupt or even collapse those ecosystems [@Haerter:2014ii; @Harcombe:2005fd; @Middelboe:2001fl; @Poisot:2011jc; @Thompson:2012ki; @Moebus:1981kp; @Flores:2013hc; @Poisot:2012fh; @Poisot071993; @Flores:2011bh; @Jover:2015ev]. The conventional approach to understanding the human microbiome is unable to capture how these communities interact and influence each other, but relationship-based network approaches allow us to capture this information. We aimed to bridge this knowledge gap by characterizing human bacterial and phage communities by their *relationships*, leveraging machine learning and graph theory techniques. We focused on characterizing the network diversity of the healthy human microbiome (including the skin and gut), so as to provide a foundation on which we and others can build further studies into the network dynamics associated with disease states.

@@

## Cohort Curation and Sample Processing
We studied the differences in microbiome phage-bacteria interaction networks across healthy human bodies by leveraging previously published sequence sets containing purified virome samples paired with bacterial metagenomes from whole metagenomic shotgun sequences. Our study contained three datasets, including a study of the impact of diet on the healthy human gut virome [@Minot:2011ez], the impact of anatomical location on the healthy human skin virome [@Hannigan:2015fz], and the viromes of monozygotic twins and their mothers [@Reyes:2010cwa; @Turnbaugh:2009ei]. The viromes associated with these datasets were subjected to virus-like particle (VLP) purification to eliminate other organismal DNA including bacteria, fungi, and humans, and allowed us to assess the actively replicating virome.

The publications we utilized were conducted over the span of five years using different methods and technologies, and therefore yielded different sequence abundances **(Supplemental Figure \ref{SequenceStats} A-B)**. The bacterial and viral sequences from those publications were quality filtered and assembled into contigs. Because contig assembly most commonly returns genome fragments, we clustered related bacteria and phage contigs by k-mer frequency and co-abundance using the CONCOCT algorithm. These clusters represent operationally defined units of related bacteria and phage genomes that we defined as operational genomic units (OGUs). These OGUs are conceptually similar to the operational taxonomic unit (OTU) and operational protein family (OPF) definitions used for grouping highly similar 16S rRNA gene and open reading frame sequences, respectively [@Schloss:2008ht]. The resulting contigs and OGUs demonstrated high sequence coverage and length **(Supplemental Figure \ref{ContigStats} - \ref{ClusterStats})**.

## Establishing a Model of Phage-Bacteria Interactions
We predicted which phage OGUs infected which bacteria using a random forest model trained on experimentally validated infectious relationships from six previous publications [@Jensen:1998vh;@Malki:2015tm;@Schwarzer:2012ez;@Kim:2012dh;@Matsuzaki:1992gw;@Edwards:2015iz]. This training set contained diverse bacteria and phages, with both broad and specific infectious ranges **(Supplemental Figure \ref{ValidationOverview} A - B)**. Phages with linear and circular genomes, as well as ssDNA and dsDNA genomes, were included in the analysis. Because this was a DNA sequencing study, RNA phages were not considered in the analysis **(Supplemental Figure \ref{ValidationOverview} C-D)**. This training set included both positive relationships (a phage infects a bacterium) and negative relationships (a phage does not infect a bacterium). This built on previous work, which focused only on positive relationships, by allowing us to validate the false positive *and* false negative rates associated with our candidate models [@Edwards:2015iz].

Four phage and bacterial genomic markers were used to predict infectious relationships between bacteria and phages: 1) genome nucleotide similarities, 2) gene amino acid sequence similarities, 3) CRISPR targeting of phages by bacterial CRISPR spacer sequences, and 4) similarity of protein families known to be associated with known protein-protein interactions [@Orchard:2014hq]. The resulting random forest model performed well, with an AUC of `r signif(genmodelper$ROC, digits = sigfig)`, a sensitivity of `r signif(genmodelper$Sens, digits = sigfig)`, and a specificity of `r signif(genmodelper$Spec, digits = sigfig)` **(Figure \ref{RocCurve} A)**. The most important predictor in the model was nucleotide similarity between genes, followed by nucleotide similarity of whole genomes **(Figure \ref{RocCurve} B)**. Protein family interactions were moderately important to the model, while CRISPRs were minimally important, likely because they were redundant with the blast information. Approximately one third of the relationships yielded no score and were therefore unknown, unable to be assigned an interaction **(Figure \ref{RocCurve} C)**.

@@

## Individuality of Microbial Networks
Skin and gut community membership and diversity are highly personal, with people remaining more similar to themselves over time compared to other people [@Grice:2009ee; @Hannigan:2015fz; @Minot:2013ih]. We determined whether this personal conservation extended to network structure using our microbiome network model. We calculated the degree of dissimilarity between each subject's network, based on phage and bacteria abundance, as well as centrality. Centrality was evaluated by first calculating the weighted eigenvector centrality of all bacteria and phages within each sample graph. Conceptually, this metric defines central phages as those that are highly abundant and infect many bacteria which themselves are abundant and infected by many other phages. Bacterial centrality was defined using the same metric. We then calculated the similarity of community networks using the weighted eigenvector centrality of all nodes between all samples. Biologically, samples with similar network structures are interpreted as having similar capacities to influence other microbes within the community, transmit genetic material (e.g. horizontal gene transfer), and maintain stability.

We used this network dissimilarity metric to test whether microbiome network structures were more similar within people over time compared to other people. We found that gut microbiome network structure clustered by person (ANOSIM p-value = `r signif(interstats[c(interstats$site %in% "DietAnosim"), "prob"], digits = sigfig)`, R = `r signif(interstats[c(interstats$site %in% "AS"), "prob"], digits = sigfig)`, **Figure \ref{intradiv} A**). We found evidence that network dissimilarity within people over the 8-10 day sampling period was less than the average dissimilarity between that person and others, although this was not statistically significant (p-value = `r signif(interstats[c(interstats$site %in% "Diet"), "prob"], digits = sigfig)`, **Figure \ref{intradiv} B**). The lack of statistical confidence was likely due to the small sample size of this dataset. Although there was evidence for gut network conservation among individuals, we found no evidence for conservation of gut network structures within families. The gut network structures were not more similar between twins and their mothers (intrafamily) compared to other families of twins and mothers (inter-family) (p-value = `r signif(interstats[c(interstats$site %in% "Skin"), "prob"], digits = sigfig)`, **Figure \ref{intradiv} C**).

Skin microbiome network structure was strongly conserved within individuals, and more adequately confirmed by our larger skin dataset (p-value = `r signif(interstats[c(interstats$site %in% "Twins"), "prob"], digits = sigfig)`, **Figure \ref{intradiv} D**). This distribution was similar when separated by anatomical sites, with most sites also being significantly more conserved within individuals **(Supplemental Figure \ref{allskin})**.

@@

We evaluated the diet-associated differences in gut network structure by quantifying how central each sample's network was on average across all microbes. We accomplished this by utilizing two common centrality metrics: degree centrality and closeness centrality. **Degree centrality**, the simplest centrality metric, was defined as the number of connections each phage made to bacteria, or each bacterium made to phages. Because this metric alone offers only minimal insight, we supplemented it with measurements of closeness centrality. **Closeness centrality** is a metric of how close each phage or bacterium is to all of the other phages and bacteria in the network. A higher closeness centrality suggests that genetic information or the effects of altered abundance would be more impactful to all other microbes in the system. A networks with higher closeness centrality also indicates an overall greater degree of connections, which suggests a greater resilience to instability. Because these values are assigned to each phage and bacterium within each network, we calculated the average connectedness and corrected for the maximum potential degree of connectedness to obtain a single value for the connectedness of each graph.

We found that gut microbiome network structures associated with high fat diets were less connected than those of low fat diets **(Figure \ref{dietnetworks} A-B)**. Tests for statistical differences were not performed due to the small sample size. High fat diets exhibited less degree centrality **(Figure \ref{dietnetworks} A)**. This suggests that bacteria were targeted by less phages and that phage tropism was more specific. High fat diets also exhibited decreased closeness centrality **(Figure \ref{dietnetworks} B)**, indicating that microbes were more distant from other microbes in the community. This would make genetic transfer and altered abundance less capable of impacting other bacteria and phages within the network.

In addition to diet, there was also an association between obesity and network structure **(Figure \ref{dietnetworks} C-D)**. The obesity-associated network demonstrated a higher degree centrality **(Figure \ref{dietnetworks} C)** but a less closeness centrality, compared to the healthy controls **(Figure \ref{dietnetworks} D)**. This suggested that the obesity network was overall less connected, having microbes further from all other microbes within the community.

@@

We quantified the average centrality of each sample again using the weighted eigenvector centrality metric. We found that intermittently moist skin sites (dynamic sites that fluctuate between being moist and dry) were significantly less connected than the more stable moist and sebaceous environments (p-value < 0.001, **Figure \ref{skinnetwork} A)**. We also found that skin sites that were protected from the environment (occluded) were much more highly connected than those that were constantly exposed to the environment or only intermittently occluded (p-value < 0.001, **Figure \ref{skinnetwork} B)**.

We supplemented this analysis by comparing the network signatures using the centrality dissimilarity approach described above. The dissimilarity between samples was a function of shared relationships, degree of centrality, and bacteria/phage abundance. When using this supplementary approach, we found that network structures significantly clustered by moisture, sebaceous, and intermittently moist status **(Figure \ref{skinnetwork} C,E)**. We also found that occluded sites were significantly different from exposed and intermittently occluded sites, but there was no difference between exposed and intermittently occluded sites **(Figure \ref{skinnetwork} D,F)**.

# Discussion
We developed and implemented a network-based analytical approach that allowed us to begin evaluating the basic properties of the human microbiome through bacteria and phage *relationships*, instead of membership or diversity. The goal of this study was to provide an initial understanding of how phage-bacteria networks differ throughout the human body, so as to provide a baseline for future studies of how microbiome networks differ in disease states. This goal was similar to the initial studies of bacteria and viral diversity across the human body, as well as other environments [@Grice:2009eea; @Findley:2013jf; @Hannigan:2015fz; @Costello:2009im, @Consortium:2012iz; @Schloss:2005hz; @Minot:2011ez]. The use of network theory offered extensive analytical opportunities which could be applied to understand complex ecological communities. We focused on utilizing metrics of connectivity to understand the extent to which communities of bacteria and phages interact (e.g. horizontal gene transfer, modulated bacterial gene expression, alterations in abundance). By accomplishing this goal, we aimed to provide an initial understanding of human microbiome networks that will power fututure studies to use similar approaches in other setting such as disease states.

We reported that, just as gut microbiome and virome composition and diversity are conserved in individuals [@Hannigan:2015fz; @Grice:2009eea; @Findley:2013jf; @Minot:2013ih], gut and skin microbiome network structures were also conserved within individuals over time. Gut network structure was not conserved among family members. These findings suggest that microbiome properties such as stability and the potential for horizontal gene transfer are personal, and potentially impacted by personal factors ranging from the body's immune system to environemntal conditions such as climate and diet. The ability of environmental conditions to shape gut and skin microbiome network structure was supported by our finding that diet and skin location were associated with altered network structures.

We found evidence that diet was sufficient to alter gut microbiome network connectivity. Although our sample size was small, we found that high fat diets appeared to be less connected than low fat diets. This suggested that high fat diets may lead to less stable communities with a decreased ability to influence each other. We supported this finding with the observation that obesity is associated with decreased network connectivity. Further work will be required to characterize these relationships with a larger cohort, but our results allow us to speculate that the food we eat may not only impact what microbes colonize our guts, but may also impact interactions.

We also found evidence that skin microbiome network structure varied by skin environments. We observed that network structure differed between environmentally exposed and occluded skin sites. The sites under greater environmental fluctuation and exposure, the exposed and intermittently exposed sites, were less connected and therefore were expected to have a higher propensity for instability. Likewise, intermittently moist sites demonstrated less connectedness than the more stable moist and sebaceous sites. Together this suggested that body sites under greater degrees of fluctuation harbored less connected, potentially less stable microbiomes.

This work represents an initial step toward understanding the microbiome through its relationships. While these findings are informative, there are certainly caveats that should be noted. *First*, while our infection classification model is advantageous over existing models, we recognize that, like most classification models, there remains opportunity for improvement. For example, such a model is only as good as its training set, and future large-scale endeavors into infectious relationships (and the associated genomes) will provide more robust training and higher model accuracy. Just as we have improved on previous modeling efforts, we expect that new and creative scoring metrics will likely be integrated into this model to further improve model performance. *Second*, while informative, this work was done retrospectively and relied on published research from as many as seven years ago. These archived datasets were limited by the technology and costs of the time, meaning the datasets are poorly powered for the statistical analysis we strive for today. While we were able to present initial observations, follow-up studies will be required to validate our observations. Despite their limitations, these results will be important for informing design and interpretation of future studies of the human microbiome.

By showing that microbiome relationship structure, and therefore the community potential for dissemination of information and influence, as well as the potential for maintaining stability, differs significantly between body sites and environmental conditions we demonstrate the diversity of relationships across the human body. We found that other environmental factors, such as diet, also influence relationship structures. This information builds on previous work by suggesting that bacteria and phages not only preferentially colonize different body sites, but also interact differently, allowing for different communication structures and capacities for maintaining stability. This work will also provide a foundation for future studies to begin evaluating how microbiome network dynamics change in disease states, and how the information can be leveraged therapeutically.

@@

OGU abundance (`Ao`) was obtained as the sum of the abundance of each contig (`Aj`) associated with that OGU. The abundance values were length corrected such that:

@@

##### Identify Bacterial CRISPRs Targeting Phages
CRISPRs were identified from bacterial genomes using the PilerCR program (v1.06) [@Edgar:2007bh]. Resulting spacer sequences were filtered to exclude spacers shorter than 20 bp and longer than 65 bp. Spacer sequences were aligned to the phage genomes using the nucleotide blast algorithm with default parameters (v2.4.0) [@Camacho:2009fc]. The mean percent identity for each matching pair was recorded for use in our classification model.

##### Detect Matching Prophages within Bacterial Genomes
Temperate bacteriophages infect and integrate into their bacterial host's genome. We detected integrated phage elements within bacterial genomes by aligning phage genomes to bacterial genomes using the nucleotide blast algorithm and a minimum e-value of 1e-10. The resulting bitscore of each alignment was recorded for use in our classification model.

@@

[//]: <> The distance between nodes (`d`) was calculated as the shortest number of edges required to be traversed to move from one node to another.

@@

[//]: <> The simple metric of degree centrality of node Vi was defined as the sum of the number of k edges Ei associated with that node, such that:

@@

The **centralization** metric was used to assess the average centrality of each sample graph `G`. Centralization was calculated by taking the sum of each vertex `Vi`'s centrality from the graph maximum centrality `Cw`, such that:

@@

## Network Relationship Dissimilarity
We assessed similarity between graphs by evaluating the shared centrality of their vertices, as has been done previously. More specifically, we calculated the dissimilarity between graphs `Gi` and `Gj` using the Bray-Curtis dissimilarity metric and Eigenvector centrality values such that:

@@

Where `Cij` is the sum of the lesser centrality values for those vertices shared between graphs, and `Ci` and `Cj` are the total number of vertices found in each graph. This allows us to calculate the dissimilarity between graphs based on the shared centrality values between the two graphs.

@@

![**Summary of Multi-Study Network Model.** *(A) ROC curve resulting from ten iterations used to create the phage - bacteria infection prediction model. (B) Importance scores associated with the metrics used in the random forest model to predict relationships between bacteria and phages. The importance score is defined as the mean decrease in accuracy of the model when a feature (e.g. Pfam) is excluded. (C) Proportions of samples included (gray) and excluded (red) in the model. Samples were excluded from the model because they did not yield any scores. Those interactions without scores were defined as not having interactions. (D) Bipartite visualization of the resulting phage-bacteria network. This network includes information from all three published studies. (E) Network diameter (measure of graph size; the greatest number of traversed vertices required between two vertices), (F) number of vertices, and (G) number of edges (relationships) for the total network (yellow) and the individual study sub-networks (diet study = red, skin study = green, twin study = orange).* \label{RocCurve}](../figures/rocCurves.pdf){ width=90% }

@@

![**Impact of Diet and Obesity on Gut Network Structure.** *(A) Quantification of average degree centrality (number of edges per node) and (B) closeness centrality (average distance from each node to every other node) of gut microbiome networks of subjects limited to exclusively high fat or low fat diets. Lines represent the mean degree of centrality for each diet. (C) Quantification of average degree centrality and (D) closeness centrality between obese and healthy adult women.*\label{dietnetworks}](../figures/dietnetworks.pdf){ width=90% }

>>>>

# Abstract
Viruses and bacteria are critical components of the human microbiome and play important roles in health and disease. Previous work has relied on studying bacteria and phages in isolation, reducing them to two separate communities. These approaches have not captured how these communities interact to share information (e.g. horizontal gene transfer) and maintain stability. We developed and implemented a network-based analytical approach to provide an initial understanding of phage-bacteria network diversity throughout the human body. We accomplished this by building a machine learning algorithm to predict which phages infected which bacteria in a given microbiome. This algorithm was applied to paired viral and bacterial metagenomic sequence sets from three previously published human cohorts. We organized the predicted interactions into networks that allowed us to evaluate the diversity of phage-bacteria connectedness across the human body. We focused on establishing a foundational understanding of microbiome network structures across the human body. We found that gut and skin network structure was person-specific, and was not conserved among cohabitating family members. High-fat diets and obesity were associated with less connected networks. There were significant differences in network structure between skin sites, with those exposed to the external environment being less connected and more prone to instability. Altogether this study characterizes the diversity of microbiome networks across the human body, and provides a baseline for future studies to investigate the role of ecological networks in disease states.

@@

Previous work has relied on studying bacterial and phage communities in isolation, reducing them to two separate communities. In reality, bacteria and phage communities are dynamic and complex, acting in concert to share genetic information and to maintain stable ecosystems, and removal of members can disrupt or even collapse those ecosystems [@Haerter:2014ii; @Harcombe:2005fd; @Middelboe:2001fl; @Poisot:2011jc; @Thompson:2012ki; @Moebus:1981kp; @Flores:2013hc; @Poisot:2012fh; @Poisot071993; @Flores:2011bh; @Jover:2015ev]. The conventional approach to understanding the human microbiome is unable to capture how these communities interact and influence each other, but relationship-based network approaches allow us to capture this information. We aimed to bridge this knowledge gap by characterizing human bacterial and phage communities by their relationships, leveraging machine learning and graph theory techniques. We focused on characterizing the network diversity of the human microbiome (including skin and gut) to provide a foundation for further studies of disease network dynamics.

@@

## Cohort Curation and Sample Processing
We studied the differences in microbiome phage-bacteria interaction networks across healthy human bodies by leveraging previously published sequence sets containing purified virome samples paired with bacterial metagenomes from whole metagenomic shotgun sequences. Our study contained three datasets, including a study of the impact of diet on the healthy human gut virome [@Minot:2011ez], the impact of anatomical location on the healthy human skin virome [@Hannigan:2015fz], and the viromes of monozygotic twins and their mothers [@Reyes:2010cwa; @Turnbaugh:2009ei]. The viromes associated with these datasets were all subjected to virus-like particle (VLP) purification to eliminate other organismal DNA including bacteria, fungi, and humans, thus allowing the original authors and ourselves to assess the actively replicating virome.

The publications we utilized were published over the span of five years using different methods and technologies, and therefore yielded different sequence abundances **(Supplemental Figure \ref{SequenceStats} A-B)**. The bacterial and viral sequences from those publications were quality filtered and assembled into contigs. Because contig assembly most commonly returns genome fragments, we clustered related bacteria and phage contigs by k-mer frequency and co-abundance using the CONCOCT algorithm. These clusters represent operationally defined units of related bacteria and phage genomes that we defined as operational genomic units (OGUs). These OGUs are conceptually similar to the operational taxonomic unit (OTU) and operational protein family (OPF) definitions used for grouping highly similar 16S rRNA gene and open reading frame sequences, respectively [@Schloss:2008ht]. The resulting contigs and OGUs demonstrated high sequence coverage and length **(Supplemental Figure \ref{ContigStats} - \ref{ClusterStats})**.

## Establishing a Model of Phage-Bacteria Interactions
We predicted which phage OGUs infected which bacteria using a random forest model trained on experimentally validated infectious relationships from six previous publications [@Jensen:1998vh;@Malki:2015tm;@Schwarzer:2012ez;@Kim:2012dh;@Matsuzaki:1992gw;@Edwards:2015iz]. This training set contained diverse bacteria and phages, with both broad and specific infectious ranges **(Supplemental Figure \ref{ValidationOverview} A - B)**. Phages with linear and circular genomes, as well as ssDNA and dsDNA genomes, were included in the analysis. Because this was a DNA sequencing study, RNA phages were not considered in the analysis **(Supplemental Figure \ref{ValidationOverview} C-D)**. This training set included both positive relationships (phage infects a bacterium) and negative relationships (phage does not infect a bacterium). This built on previous work, which focused only on positive relationships, by allowing us to validate the false positive *and* false negative rates associated with our candidate models [@Edwards:2015iz].

Four phage and bacterial genomic markers were used to predict infectious relationships between bacteria and phages: 1) genome nucleotide similarities, 2) gene amino acid sequence similarities, 3) CRISPR targeting of phages by bacterial CRISPR spacer sequences, and 4) similarity of protein families known to be associated with known protein-protein interactions [@Orchard:2014hq]. The resulting random forest model performed well, with an AUC of `r signif(genmodelper$ROC, digits = sigfig)`, a sensitivity of `r signif(genmodelper$Sens, digits = sigfig)`, and a specificity of `r signif(genmodelper$Spec, digits = sigfig)` **(Figure \ref{RocCurve} A)**. The most important predictor in the model was nucleotide similarity between genes, followed by nucleotide similarity of whole genomes **(Figure \ref{RocCurve} B)**. Protein family interactions were moderately important to the model, while CRISPRs were minimally important, likely because they were redundant with the BLAST information. Approximately one third of the relationships yielded no score and were therefore unknown, unable to be assigned an interaction **(Figure \ref{RocCurve} C)**.

@@

## Individuality of Microbial Networks
Skin and gut community membership and diversity are highly personal, with people remaining more similar to themselves over time compared to other people [@Grice:2009ee; @Hannigan:2015fz; @Minot:2013ih]. We determined whether this personal conservation extended to network structure using our microbiome network model. We calculated the degree of dissimilarity between each subject's network based on phage and bacteria abundance, as well as centrality. Centrality was evaluated by first calculating the weighted eigenvector centrality of all bacteria and phages within each sample graph. Conceptually, this metric defines central phages as those that are highly abundant and infect many bacteria which themselves are abundant and infected by many other phages. Bacterial centrality was defined using the same metric. We then calculated the similarity of community networks using the weighted eigenvector centrality of all nodes between all samples. Biologically, samples with similar network structures are interpreted as having similar capacities to influence other microbes within the community, transmit genetic material (e.g. horizontal gene transfer), and maintain stability.

We used this network dissimilarity metric to test whether microbiome network structures were more similar within people over time compared to other people. We found that gut microbiome network structure clustered by person (ANOSIM p-value = `r signif(interstats[c(interstats$site %in% "DietAnosim"), "prob"], digits = sigfig)`, R = `r signif(interstats[c(interstats$site %in% "AS"), "prob"], digits = sigfig)`, **Figure \ref{intradiv} A**). We found evidence that network dissimilarity within people over the 8-10 day sampling period was less than the average dissimilarity between that person and others, although this was not statistically significant (p-value = `r signif(interstats[c(interstats$site %in% "Diet"), "prob"], digits = sigfig)`, **Figure \ref{intradiv} B**). The lack of statistical confidence was likely due to the small sample size of this dataset. Although there was evidence for gut network conservation among individuals, we found no evidence for conservation of gut network structures within families. The gut network structures were not more similar between twins and their mothers (intrafamily) compared to other families of twins and mothers (inter-family) (p-value = `r signif(interstats[c(interstats$site %in% "Twins"), "prob"], digits = sigfig)`, **Figure \ref{intradiv} C**).

Skin microbiome network structure was strongly conserved within individuals, and more adequately confirmed by our larger skin dataset (p-value = `r signif(interstats[c(interstats$site %in% "Skin"), "prob"], digits = sigfig)`, **Figure \ref{intradiv} D**). This distribution was similar when separated by anatomical sites, with most sites also being significantly more conserved within individuals **(Supplemental Figure \ref{allskin})**.

@@

We evaluated the diet-associated differences in gut network structure by quantifying how central each sample's network was on average. We accomplished this by utilizing two common centrality metrics: degree centrality and closeness centrality. Degree centrality, the simplest centrality metric, was defined as the number of connections each phage made to bacteria, or each bacterium made to phages. Because this metric alone offers only minimal insight, we supplemented it with measurements of closeness centrality. Closeness centrality is a metric of how close each phage or bacterium is to all of the other phages and bacteria in the network. A higher closeness centrality suggests that genetic information or the effects of altered abundance would be more impactful to all other microbes in the system. A network with higher average closeness centrality also indicates an overall greater degree of connections, which suggests a greater resilience to instability. Because these values are assigned to each phage and bacterium within each network, we calculated the average connectedness and corrected for the maximum potential degree of connectedness to obtain a single value for the connectedness of each graph.

We found that gut microbiome network structures associated with high-fat diets were less connected than those of low-fat diets **(Figure \ref{dietnetworks} A-B)**. Tests for statistical differences were not performed due to the small sample size. High-fat diets exhibited less degree centrality **(Figure \ref{dietnetworks} A)**. This finding suggests that bacteria were targeted by less phages and that phage tropism was more specific. High-fat diets also exhibited decreased closeness centrality **(Figure \ref{dietnetworks} B)**, indicating that microbes were more distant from other microbes in the community. This would make genetic transfer and altered abundance less capable of impacting other bacteria and phages within the network.

In addition to diet, we also observed an association between obesity and network structure **(Figure \ref{dietnetworks} C-D)**. The obesity-associated network demonstrated a higher degree centrality **(Figure \ref{dietnetworks} C)** but less closeness centrality compared to the healthy controls **(Figure \ref{dietnetworks} D)**. These results suggest that the obesity network was overall less connected, having microbes further from all other microbes within the community.

@@

We quantified the average centrality of each sample using the weighted eigenvector centrality metric. We found that intermittently moist skin sites (dynamic sites that fluctuate between being moist and dry) were significantly less connected than the more stable moist and sebaceous environments (p-value < 0.001, **Figure \ref{skinnetwork} A)**. We also found that skin sites that were protected from the environment (occluded) were much more highly connected than those that were constantly exposed to the environment or only intermittently occluded (p-value < 0.001, **Figure \ref{skinnetwork} B)**.

We supplemented this analysis by comparing the network signatures using the centrality dissimilarity approach described above. The dissimilarity between samples was a function of shared relationships, degree of centrality, and bacteria/phage abundance. When using this supplementary approach, we found that network structures significantly clustered by moisture, sebaceous, and intermittently moist status **(Figure \ref{skinnetwork} C,E)**. We also found that occluded sites were significantly different from exposed and intermittently occluded sites, but there was no difference between exposed and intermittently occluded sites **(Figure \ref{skinnetwork} D,F)**. These findings provide further support that skin microbiome network structure differs significantly between skin sites.

# Discussion
We developed and implemented a network-based analytical approach which we used to begin evaluating the basic properties of the human microbiome through bacteria and phage *relationships*, instead of membership or diversity. The goal of this study was to provide an initial understanding of how phage-bacteria networks differ throughout the human body, so as to provide a baseline for future studies of how microbiome networks differ in disease states. This goal of providing a baseline understanding of the human microbiome was similar to the goals of initial studies of bacteria and viral diversity across the human body, as well as other environments [@Grice:2009eea; @Findley:2013jf; @Hannigan:2015fz; @Costello:2009im, @Consortium:2012iz; @Schloss:2005hz; @Minot:2011ez]. Through the use of network theory, we leveraged extensive analytical opportunities which could be applied to understand complex ecological communities. We focused on utilizing metrics of connectivity to understand the extent to which communities of bacteria and phages interact (e.g. horizontal gene transfer, modulated bacterial gene expression, alterations in abundance). By pursuing this goal, we aimed to provide an initial understanding of human microbiome networks that will power future studies to use similar approaches in other setting such as disease states.

We reported that, just as gut microbiome and virome composition and diversity are conserved in individuals [@Hannigan:2015fz; @Grice:2009eea; @Findley:2013jf; @Minot:2013ih], gut and skin microbiome network structures were also conserved within individuals over time. Gut network structure was not conserved among family members. These findings suggest that microbiome properties such as stability and the potential for horizontal gene transfer are personal, and potentially impacted by personal factors ranging from the body's immune system to environmental conditions such as climate and diet. The ability of environmental conditions to shape gut and skin microbiome network structure was supported by our finding that diet and skin location were associated with altered network structures.

We found evidence that diet was sufficient to alter gut microbiome network connectivity. Although our sample size was small, we found evidence that high-fat diets were less connected than low-fat diets. This suggested that high-fat diets may lead to less stable communities with a decreased ability to influence each other. We supported this finding with the observation that obesity was associated with decreased network connectivity. Our results allow us to speculate that the food we eat may not only impact what microbes colonize our guts, but may also impact their interactions. Further work will be required to characterize these relationships with a larger cohort.

In addition to diet, we found evidence that skin microbiome network structure varied by skin environment. We observed that network structure differed between environmentally exposed and occluded skin sites. The sites under greater environmental fluctuation and exposure, the exposed and intermittently exposed sites, were less connected and therefore were expected to have a higher propensity for instability. Likewise, intermittently moist sites demonstrated less connectedness than the more stable moist and sebaceous sites. Thus body sites under greater degrees of fluctuation harbored less connected, potentially less stable microbiomes.

Our results represent a step toward understanding the microbiome through interspecies relationships. While these findings are informative, there are certainly caveats that should be noted. *First*, while our infection classification model is advantageous over existing models, we recognize that, like most classification models, there remains opportunity for improvement. For example, such a model is only as good as its training set, and future large-scale endeavors into infectious relationships (and the associated genomes) will provide more robust training and higher model accuracy. Just as we have improved on previous modeling efforts, we expect that new and creative scoring metrics will likely be integrated into this model to further improve model performance. *Second*, while informative, this work was done retrospectively and relied on published research from as many as seven years ago. These archived datasets were limited by the technology and costs of the time, meaning the datasets are poorly powered for the statistical analysis we strive for today. While we were able to present initial conclusions, follow-up studies will be required to validate our findings. Despite their limitations, these methods and results will be important for informing design and interpretation of future studies of the human microbiome.

We demonstrated the diversity of relationships across the human body by showing that microbiome relationship structure, and therefore the potential for genetic transfer and stability maintenance, differs significantly between body sites and environmental conditions. We found that other environmental factors, such as diet, also influence relationship structures. This information builds on previous work by suggesting that bacteria and phages not only preferentially colonize different body sites, but also interact differently, allowing for different communication structures and capacities for maintaining stability. Our results will also provide a foundation for future studies to begin evaluating how microbiome network dynamics change in disease states, and how the information can be leveraged therapeutically.

@@

OGU abundance ($A_{O}$) was obtained as the sum of the abundance of each contig ($A_{j}$) associated with that OGU. The abundance values were length corrected such that:

@@

##### Identify Bacterial CRISPRs Targeting Phages
CRISPRs were identified from bacterial genomes using the PilerCR program (v1.06) [@Edgar:2007bh]. Resulting spacer sequences were filtered to exclude spacers shorter than 20 bp and longer than 65 bp. Spacer sequences were aligned to the phage genomes using the nucleotide BLAST algorithm with default parameters (v2.4.0) [@Camacho:2009fc]. The mean percent identity for each matching pair was recorded for use in our classification model.

##### Detect Matching Prophages within Bacterial Genomes
Temperate bacteriophages infect and integrate into their bacterial host's genome. We detected integrated phage elements within bacterial genomes by aligning phage genomes to bacterial genomes using the nucleotide BLAST algorithm and a minimum e-value of 1e-10. The resulting bitscore of each alignment was recorded for use in our classification model.

@@

The distance between nodes (`d`) was calculated as the shortest number of edges required to be traversed to move from one node to another.

@@

@@

The **centralization** metric was used to assess the average centrality of each sample graph $\mathbf{G}$. Centralization was calculated by taking the sum of each vertex $V_{i}$'s centrality from the graph maximum centrality $C_{w}$, such that:

@@

## Network Relationship Dissimilarity
We assessed similarity between graphs by evaluating the shared centrality of their vertices, as has been done previously. More specifically, we calculated the dissimilarity between graphs $G_{i}$ and $G_{j}$ using the Bray-Curtis dissimilarity metric and Eigenvector centrality values such that:

@@

Where $C_{ij}$ is the sum of the lesser centrality values for those vertices shared between graphs, and $C_{i}$ and $C_{j}$ are the total number of vertices found in each graph. This allows us to calculate the dissimilarity between graphs based on the shared centrality values between the two graphs.

@@

![**Summary of Multi-Study Network Model.** *(A) Average ROC curve resulting from ten iterations used to create the phage - bacteria infection prediction model. (B) Importance scores associated with the metrics used in the random forest model to predict relationships between bacteria and phages. The importance score is defined as the mean decrease in accuracy of the model when a feature (e.g. Pfam) is excluded. (C) Proportions of samples included (gray) and excluded (red) in the model. Samples were excluded from the model because they did not yield any scores. Those interactions without scores were defined as not having interactions. (D) Bipartite visualization of the resulting phage-bacteria network. This network includes information from all three published studies. (E) Network diameter (measure of graph size; the greatest number of traversed vertices required between two vertices), (F) number of vertices, and (G) number of edges (relationships) for the total network (yellow) and the individual study sub-networks (diet study = red, skin study = green, twin study = orange).* \label{RocCurve}](../figures/rocCurves.pdf){ width=90% }

@@

![**Impact of Diet and Obesity on Gut Network Structure.** *(A) Quantification of average degree centrality (number of edges per node) and (B) closeness centrality (average distance from each node to every other node) of gut microbiome networks of subjects limited to exclusively high-fat or low-fat diets. Lines represent the mean degree of centrality for each diet. (C) Quantification of average degree centrality and (D) closeness centrality between obese and healthy adult women.*\label{dietnetworks}](../figures/dietnetworks.pdf){ width=90% }
